---
title: "data_wrangling"
author: "Marie Rivers"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```

Look at `us_rent_income` (built-in to the `tidyr` package). Make a tidy version. Then from your tidy version, get it back to the original structure
```{r}
us_rent_wide <- us_rent_income %>% 
  select(-moe) %>% 
  pivot_wider(names_from = variable, values_from = estimate)
```

## Practice some tidyverse functions for data wrangling

- `filter()`
- `select()`
- `mitage()`
- `case_when()`
- `group_by() %>% summarize()`
- piping things together in sequence

### dplyr::filter() to keep or exclude rows based on our conditions

1. Exact match (single condition)  
Only keep observations (rows) where the island is Biscoe
```{r}
ex_1 <- filter(.data = penguins, island == "Biscoe") # island MATCHES Biscoe

unique(ex_1$island)
```
2. "and" statement within filter  
Only keep observations if the penguin species is Chinstrap AND the year is 2008
```{r}
ex_2 <- penguins %>% 
  filter(species == "Chinstrap", year == 2008)
# or
ex_2_alt <- penguins %>% 
  filter(species - "Chinstrap" & year == 2008)
# or
#ex_2_alt2 <- penguins %>% 
  filter(species == "Chinstrap" %>% 
           filter(year == "2008"))
```
3. Exclude Gentoos  
Keep observations if the penguin species is NOT "Gentoo"
```{r}
ex_3 <- penguins %>% 
  filter(species != "Gentoo")
```
Exercies: Create a subset from `penguins` that only contains observations from Adelie penguins with a flipper length greater than or equal to 185 mm.
```{r}
practice_1 <- penguins %>% 
  filter(species == "Adelie", flipper_length_mm >= 185)
```

4. The %in% operator  
Use the "in" `%in%` operator to look for *any* matches within a vector, and return TRUE if the value does exist for our row.  
We want to keep observations wehre the island is Dream or Torgersen
```{r}
ex_4 <- penguins %>% 
  filter(island %in% c("Dream", "Torgersen"))
# it is VERY important to use %in% in this example so that in each row you only look for "Dream" or "Torgerson" instead of instances where "Dream" is followed by "Torgersen"
```

What would actually be happening if in the code above if you used `==` instead of `%in`

## don't you double equal signs `==` with a vector becuase then filtering tries to match the order of that vector (ie instances where "Dream" is followed by "Torgersen")

```{r}
ex_4_only_biscoe <- penguins %>% 
  filter(!island %in% c("Dream", "Torgersen"))
```

### select() to select or exclude COLUMNS  
5. Keep a single column by name  
Only want to keep the species column

```{r}
ex_5 <- penguins %>%  select(species)
# or
ex_5_alt <- select(.data = penguins, species)
```

6. Keep multiple columns by name  
Create a subset that contains species, bill_length_mm through flipper_length_mm, and year  
```{r}
ex_6 <- penguins %>% 
  select(species, bill_length_mm:flipper_length_mm, year)
names(ex_6) # `names()` returns column names
```

7. Choose or exclude columns based on position
```{r}
ex_7 <- penguins %>% 
  select(-4) # select everything except for column 4
```

8. A piped sequence  
Write a piped sequence where you first only keep columns for species, body_mass_gm, and island, and second: filter to only include observations if body_mass_g is greater than 3600.
```{r}
ex_8 <- penguins %>% 
  select(species, body_mass_g, island) %>% 
  filter(body_mass_g >= 3600)
```

### `mutate()` to add or update existing columns, while keeping the ones already there  
9. Add a column that has body mass coverted to kg  
```{r}
ex_9 <- penguins %>% 
  mutate(body_mass_kg = body_mass_g / 1000,
         flipper_length_m = flipper_length_mm /1000) 

# start with name of new column. If you give it the name of a column that already exists within the data frame it will overwrite that column

# use a single equal sign `=` because we aren't asking "does this match?"
# you would need quotes abour the new column name if there was a space in the name, but otherwise you don't need quotes
```

10. Mutate to update an existing column
(when in doubt create a new column)
```{r}
ex_10 <- penguins %>% 
  mutate(island = str_replace(island, pattern = "sen", replacement = "HOORAY!"))
# this can be helpful to fix typos or data entry errors

ex_10_alt <- penguins %>% 
  mutate(island_hooray = str_replace(island, pattern = "sen", replacement = "HOORAY!"))
```

11. Update the class of the bill depth column to character
```{r}
ex_11 <- penguins %>% 
  mutate(bill_depth_mm = as.character(bill_depth_mm))
# mutate(bill_depth_mm = is.numeric(bill_depth_mm) is ASKING "is this numeric?" and the output would be TRUE or FALSE
class(ex_11$bill_depth_mm)
```
### mutate for adding new columns or changing exisitng columns

### case_when(): your favorite if-else  
12. Create a new column that contains "small" if penguin flipper length is less than 185 mm, "medium" otherwise
```{r}
ex_12 <- penguins %>% 
  mutate(binned_size = case_when(
    flipper_length_mm < 185 ~ "small", # the `~` symbol means "then"
    flipper_length_mm >= 185 ~ "medium"
  ))
```
Task: Starting from penguins, create a new column "bird_bill_bin" that contains:  
- If bill length < 38mm: "short bill"
- If bill length >= 38mm < 43mm: "medium bill"
- If bill length >= 43mm: "long bill"
```{r}
# `case_when()`
bill_bin <- penguins %>% 
  mutate(bird_bill_bin = case_when(
    bill_length_mm < 38 ~ "short bill",
    bill_length_mm >= 38 & bill_length_mm < 43 ~ "medium bill", # or could use the "between" function 
    bill_length_mm >= 43 ~ "long bill"
  ))
# to populate values where "anything else is true" the last line would be TRUE ~ "long bill" (this includes NA values)
```
### group_by() %>% summarize() to find summay values by group, and return them in a nice summary table
```{r}
mean(penguins$bill_depth_mm, na.rm = TRUE)
```
13. Use `group_by()` to say there are groups within the species column
Create a summary table by given groups
```{r}
# group_by() doesn't actually change the structure or content of your data frame
ex_13 <- penguins %>% 
  group_by(species, island) %>% # group by species and island 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE)) # give column names to the output table
ex_13
```
```{r}
# add a new column with the mean flipper lenght of the species in the observation
penguins %>%
  group_by(species) %>% 
  mutate(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE))
```

## Practice  
In a single piped sequence:
- only keep penguins that have a body mass > 3500 g
- only keep columns for species, flipper_lenght_mm, and sex
- create a summary table that contains the mean, standard deviation, max, and min value of flipper length by species and sex (as the grouping variables)
```{r}
ex_14 <- penguins %>% 
  filter(body_mass_g > 3500) %>%
  select(species, flipper_length_mm, sex) %>% 
  group_by(species, sex) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE),
            max_flipper_length = max(flipper_length_mm),
            min_flipper_length = min(flipper_length_mm))
ex_14
```

